---
title: " R Bookclub"
subtitle: "Chapters 2 & 3"
author: Shu Asai
date: "04/03/2019"
output:
    beamer_presentation:
    widescreen: true
    
classoption: "aspectratio=169" 
---

```{r setup, include=FALSE}
library(arsenal)
library(tidyverse)
knitr::opts_chunk$set(echo=TRUE, fig.align = "center", fig.height = 3, fig.width = 5, fig.align = "center",comment=NA)
options(stringsAsFactors = FALSE)
```

# Chapter 2: Workflow basics 

## R as a calculator {.build}
```{r}
2+4
```

```{r}
pi
(pi+1)/2
```


## Objects & Naming {.build}
Use   **' <- '**   to attribute value to objects  
<br />


```{r}
floors_harwick <- 8
```

```{r}
floors_gonda <- 19
```

Exercise:  
```{r}
floor_diff <- floors_gonda - floors_harwick
```

```{r}
floor_diff
```

  

## Functions {.build}  
Provide input arguments, Obtain output  

Example: **seq()**  

  - Two arguments needed: seq(from, to)
  - R produces a **seq**uence of integers from provided arguments
  
```{r}
seq(10,20)
```

Function documentation is found by typing "?" + function name in the Console: 
```{r, eval=FALSE, echo=TRUE}
?seq
```


## Exercises {.build}  
Find the error in my code
```{r, eval=FALSE, echo=TRUE}
ggplot(dota = mpg) +
  geom_point(mapping = aes(x=displ, y=hwy))

```

```{r}
ggplot(data = mpg) +
  geom_point(mapping = aes(x=displ, y=hwy))
```


## Exercises cont. {.build}

```{r,eval=F, echo=TRUE}
fliter(mpg, cyl=8)
```


```{r}
filter(mpg, cyl==8)
```

## Exercises cont. {.build}

```{r,eval=FALSE, echo=TRUE}
filter(diamond, carat>3)
```

```{r}

filter(diamonds, carat>3)
```

## Keyboard Tips & Shortcuts {.build}
> - Snake Case (Examples: "floors_harwick" , "cohort_A" , "scatter_MCHS")  

      * Descriptive and readable object names
      * Avoid camel case and periods
  
> - Use **(Alt) (-)** as a keyboard shortcut for **<-**  
  
> - Using "Tab" to explore objects  
  
> - RStudio Environment & History: (Cmd) (Ctrl) (-) (up arrow)   
  
> - Shortcut Cheatsheet:  (Alt) (Shift) (K)   


# Chapter 3: Transformations with dplyr

## Variable Types {.build}
* *int* (Integers)  
  
* *dbl* (Doubles or real numbers)  
  
* *chr* (Character vectors or strings)  
  
* *dttm* (Date / Times)



## ?dplyr {.build}
  **What** is it?   
  
    dplyr is an R package containing functions useful to transform and manipulate data.  
  **Why** dplyr?  
  
    Intuitive, readable, and fast!

  Five key dplyr functions:   
   
    1. filter
    2. arrange
    3. select  
    4. mutate  
    5. summarize
  

    
## dplyr Function Workflow {.build}

Load in the dplyr package
```{r}
library(dplyr)
```


> 1. Provide the function a dataframe
> 2. Describe what to do to the dataframe with column specifications
> 3. Output to a new dataframe



## The Data {.build}
```{r, eval=TRUE, echo=F, include=FALSE}
library(nycflights13)
library(dplyr)
```

```{r}
flights
```

# 'filter': Subset Rows


## 'filter' workflow {.build}
* Provide the function a dataframe   

```{r, eval=F,echo=T}
filter(flights)
```


* Describe what to do to the dataframe with column specifications

```{r, eval=F, echo=T}
filter(flights, month == 1 , day == 1)
```


* Output to a new dataframe  

```{r}
(flights_1_1 <- filter(flights,  month == 1 , day == 1))
```

## Useful Operators{.build}
Use operators to specify your "filter" statement. 

  > 1. "and" statement: **&** 
  > 2. "or" statement: **|** 
  > 3. "not" statement: **!=** 
  > 4. extended "or" statement: **%in%** 

<br />

Find all flights operated by United ("UA"), American ("AA"), or Delta ("DL")
  
```{r,eval=F, echo=T}
filter(flights, carrier %in% c( "UA", "AA",  "DL" ))
```

## More Exercises {.build}
Find flights that had an arrival delay of two or more hours: 
```{r, eval=FALSE}
filter(flights, arr_delay >=2)
```

Find flights that flew to "IAH" or "HOU"
```{r, eval=FALSE}
filter(flights, dest == "IAH" | dest == "HOU")
filter(flights, dest %in% c("IAH","HOU"))
```

Find flights that were delayed by at least an hour, but made up over 30 minutes. 
```{r, eval=F}
filter(flights, dep_delay >= 1 & arr_delay < -30)
```


## Exercises cont. {.build}
The *between( column, min, max)* function is also a useful dplyr verb.  

```{r,eval=FALSE}
filter( dataframe, between(column, min, max) )
```

How can we use it in this example?

Find flights that departed between July, August, and September.

```{r,eval=F}
filter(flights, between(month,7,9))
```
```{r, echo=FALSE}
filter(flights, between(month,7,9))[1:3,]
```

## Exercises cont. {.build}
Which flights are missing a "dep_time"?

```{r, eval=F}
filter(flights, is.na(dep_time))
```

```{r, echo=FALSE}
filter(flights, is.na(dep_time))[1:3,]
```

What could these rows represent?


# 'arrange': Order your Data 


## 'arrange' {.build}

Order dataset by provided columns in ascending order

```{r, eval=T, echo=T}
(flight_month <- arrange(flights, month))
```

Default: ascending order

## 'arrange' cont. {.build}

Arranging the data in the opposite direction (descending order):  
```{r}
(flight_year_month <- arrange(flights, desc(month)))
```

## 'arrange': cont. {.build}
  Multiple arguments in the arrange statement.  

```{r, eval=F}
flight_year_dep <- arrange(flights, year, dep_time)
```

```{r, echo=F}
(flight_year_dep <- arrange(flights, year, dep_time))[1:4,]
```

  *What happens to NA values in arrange() statements?*
  
## Handling NAs and Missing Values {.build}
> To determine if a value is missing/NA, use the 'is.na()' function:
  
```{r}
x <- NA
is.na(x)
```
  
> The following result in NA:
  
```{r,eval=F, echo=TRUE}
NA > 5
10 == NA
NA + 10
NA / 2
```

> *Rule-* if the calculation involves NA, result will almost always be NA. 
  


## NAs: Exceptions {.build}

```{r}
NA ^ 0
```


```{r}
NA | TRUE
```



## Exercises {.build}
How can we sort missing values to the top of the dataset?

```{r,eval=F}
arrange(flights, desc(is.na(dep_time)))
```

Find flights that were most delayed, then sort by those that left earliest
```{r,eval=FALSE}
arrange(flights, desc(dep_delayed), departure)
```

Find the longest flights
```{r,eval=F}
arrange(flights, desc(air_time))
```


# 'select': Select Columns of Data

## 'select' {.build}

```{r}
flights_ymd <- select(flights, year, month, day)
flights_ymd
```

## 'select' cont. {.build}

Selecting all columns between 'carrier' and 'origin' 
```{r}
select(flights, carrier:origin)
```

## 'select' cont. {.build}

Selecting all columns except those from 'carrier' to 'origin'

```{r}
select(flights, -(carrier : origin))
```



## 'select' helper statements {.build}

> starts_with("x")

    selects columns that begin with "x"
    
> ends_with("type")

    selects columns that end with "type"
    
> contains("gene")

    selects columns that contain "gene"
    
> num_range("x", 1:3)

    selects columns titled, "x1", "x2", and "x3" 
    

## Exercises {.build}

What happens if you refer to the same column multiple times in a 'select' call?
```{r, eval=FALSE}
select(flights, year, year)
```


```{r, echo=FALSE}
select(flights, year, year)[1:3,]
```

## Exercises cont. {.build}

What does the 'one_of' function do? How could it be useful in a select statement?

Given a vector of characters, 'one_of' finds column names that match in the vector.  

```{r}
variables <- c("year", "month", "day","arr_time")
```


```{r,eval=FALSE}
select(flights, one_of(variables))
```

```{r,echo=FALSE}
select(flights, one_of(variables))[1:3,]
```


## Column Manipulation {.build}
Use 'rename' to change column names

Syntax: rename(data, *new_column_name* = old_column_name)
  
```{r}
rename(flights, depart_delay = dep_delay)
```


## Column Manipulation cont. {.build}

Use 'select' and 'everything' together to rearrange columns
```{r}
select(flights, time_hour, air_time, everything())
```

# mutate: Create Columns

## 'mutate' {.build}

Create and append columns to exisiting data with 'mutate'

Syntax: 

<center> mutate(data, *new_column_name* = column manipulation ) </center> 

## 'mutate' example {.build}
```{r}
flights_A <- select(flights, year : day, arr_delay, starts_with("dep"))
```

```{r, echo=FALSE}
flights_A[1:3,]
```

```{r, eval=F, echo=T}
mutate(flights_A, 
       gain = arr_delay - dep_delay)
```

```{r, echo=FALSE}
mutate(flights_A, gain = arr_delay - dep_delay)[1:3,]
```

## 'mutate' Multiple Columns {.build}

```{r, eval=FALSE}
mutate(flights_A, 
       gain = arr_delay - dep_delay,
       arr_dep_diff = arr_delay - dep_time)
```

```{r, echo=FALSE}
mutate(flights_A, 
       gain = arr_delay - dep_delay,
       arr_dep_diff = arr_delay - dep_time)[1:5,]
```

## 'transmute' {.build}

'transmute' is the same as 'mutate', but will instead only keep the new variables: 

```{r, echo=TRUE}
transmute(flights_A, 
       gain = arr_delay - dep_delay,
       arr_dep_diff = arr_delay - dep_time)
```



## Useful Functions When Mutating {.build}

> / , + , -,  * , ^
    
> sum(), mean()
  
> %/% (integer division)

```{r}
4 %/% 2
```
  
> %% (remainder)

```{r}
4 %% 2
```

## Useful Functions When Mutating cont. {.build}

  Logs

  * Natural logarithms
```{r, eval=FALSE}
log()
```
  
  * Binary logarithms (base 2)
```{r, eval=FALSE}
log2()
```
  * Common logarithms (base 10)
```{r, eval=FALSE}
log10()
```

## Useful Functions cont : Offsets {.build} 
  'lead' & 'lag'
  
```{r}
(tn <- 1:10)
(lead(tn))
(lag(tn))
```
  
## Useful Functions cont : Aggregates {.build} 
Cumulative and rolling calculations 

  What is the sum of all previous values (including current value)?
```{r}
cumsum(tn)
```
  
  What is the mean of all previous values (including current value)?
```{r}
cummean(tn)
```
  
  others: 'cummin' , 'cummax', 'cumprod'

  
## Useful Functions cont : Comparisons {.build} 
Logical comparisons

```{r,eval=FALSE, echo=TRUE}
<
>
<=
>= 
!= 
==
```


## Exercises {.build}

Find the 10 most delayed flights using mutate and the 'min_rank' function.  

**Note:** 'min_rank' assigns a number equal to a number of elements less than that value plus one.
```{r}
mr <- c(4, 2, 7, 7, 7 , 0, 10)
min_rank(mr)
min_rank(desc(mr))
```

## Exercises cont. {.build}

Find the 10 most delayed flights using mutate and the 'min_rank' function.  

```{r}
flights_delay <- mutate(flights, delay_rank = min_rank(desc(dep_delay)))
```

```{r}
flights_delay <- select(flights_delay, delay_rank, dep_delay, everything())
```

```{r}
arrange(flights_delay, delay_rank)[1:5,]
```



## Exercises cont. {.build}



# 'summarize': Aggregate Calculations

## 'summarize' {.build}
  Collapses data into a single row: 

```{r}
summarize(flights, avg_delay = mean(dep_delay, na.rm = TRUE))
```

  'summarize' is not too useful on its own.  
  It is often used in tandem with 

<center> **Pipes and 'group_by'** </center> 

# Pipes and 'group_by'

## Grouped Summaries with "group_by" {.build}

Implement *group_by* to use dplyr functions on a *grouped* dataframe

Example: 
```{r}
by_day <- group_by(flights, year, month, day)
```
The above code does nothing to the structure of the data. The data is merely coerced into a grouped dataframe.

```{r}
summarize(by_day, delay = mean(dep_delay, na.rm = TRUE))[1:3,]
```

## A note on 'na.rm' {.build}
Recall, *most* calculations that include 'NA' values will result in an 'NA'.

```{r}
mean(c(2,3,4,NA, 6))
```


Omit the 'NA' values in the calculation with use the option *na.rm=TRUE*

```{r}
mean(c(2,3,4,NA,6), na.rm = TRUE)
```
## Ungrouping {.build}

Data can be ungrouped at any time 
```{r}
by_day <- group_by(flights, year, month, day)
```

```{r}
orig_data <- ungroup(by_day)
```

## Piping with '%>%'  {.build}

  Pipes allow multiple 'dplyr' functions to be used consecutively. 

  For example, you may want to take your original data, and then: 
  
  1. Select specific columns
  
  2. Filter for observations 
  
  3. Create (mutate) another column

<br />

  <center>  Data **%>%** Select columns **%>%** Filter observations **%>%** Create a column </center>


## Pipe Framework {.build}
```{r,eval=FALSE}
flight_a <- flights %>% select(year, air_time, dest) %>% mutate(air_time_hours = air_time/60)
```

> 1. Provide a dataset 
> 2. Pipe ( %>% )
> 3. Input desired 'dplyr' function
> 4. Repeat steps 2 - 3 as needed
> 5. Output into a new object



```{r, echo=FALSE}
(flights %>% select(year, air_time, dest) %>% mutate(air_time_hours = air_time/60))[1:3,]
```


## Pipes and 'group_by' together {.build}

> Using both pipes and 'group_by' allows data manipulation to groups of data. 

> Consider the question: What was the average departure delay in each month? 

> 'group_by' month, use the 'summarize' function

```{r, eval=FALSE}
flights %>% group_by(month) %>% summarize(avg_delay = mean(dep_delay,na.rm=TRUE))
```
```{r, echo=F}
(flights %>% group_by(month) %>% summarize(avg_delay = mean(dep_delay,na.rm=TRUE)))[1:4,]
```

## Useful Summary Functions {.build}

```{r, eval=FALSE}
flights %>% group_by(month) %>% summarize()
```

> **Location:** mean(x), median(x)

> **Spread:** sd(x), IQR(x), mad(x) *(median absolute deviation)* 

> **Rank:** min(x), quantile( x, 0.5) , max(x)

> **Position:** first(x) , nth(x, 5), last(x)

> **Counts:** n(), sum(is.na(x)), sum(!(is.na(x))), n_distinct(x)

Find the sum of 'NAs' in the origin column in each month:  
```{r, eval=FALSE}
flights %>% group_by(month) %>% summarize(na_origin = sum(is.na(origin)))
```


## Exercises {.build}


Provide another approach to attain output from the following: 

```{r,eval=F}
not_cancelled %>% count(dest)
```
Essentially: Find all flights that are not cancelled.  

One solution: 
```{r,eval=F}
not_cancelled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))
```

Other solutions?

If a flight never departs, then it won't arrive. But a flight could also depart and not arrive (crashes, lost flights). We could use 'arr_delay' as a proxy to define cancelled flights.  


## Exercises cont. {.build}

For each plane, count the number of flights before the first delay of greater than one hour. 

What dplyr verbs do we want to use?

```{r,eval=F}
flights %>%
  arrange(tailnum, year, month, day) %>%
  group_by(tailnum) %>%
  mutate(delay_gt1hr = dep_delay > 60) %>%
  mutate(before_delay = cumsum(delay_gt1hr)) %>%
  filter(before_delay < 1) %>%
  count(sort = TRUE)
```



## Exercises {.build}

For each destination, compute the total minutes of delay.  

For each flight, compute the proportion of the total delay for its destination. 

```{r, eval=F}
flights %>%
  filter(!is.na(arr_delay), arr_delay > 0) %>%
  group_by(dest) %>%
  mutate(
    arr_delay_total = sum(arr_delay),
    arr_delay_prop = arr_delay / arr_delay_total) %>% 
  ungroup()
```


## Exercises cont. {.build}

Find destinations that are flown by at least two carriers. 

```{r, eval=FALSE}
dest_2carriers <- flights %>%
  # keep only unique carrier,dest pairs
  select(dest, carrier) %>%
  group_by(dest, carrier) %>%
  filter(row_number() == 1) %>%
  # count carriers by destination
  group_by(dest) %>%
  mutate(n_carrier = n_distinct(carrier)) %>%
  filter(n_carrier >= 2)
```

